---
title: '3) The Basics'
author: "Jasmine Hughes"
date: "12/01/2023"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# R as a calculator

Write some example expressions for each of the math operations below:

```{r computing,echo=TRUE,results='markup'}
# add numbers using +

# multiply numbers using *

# powers: two ways! `**` or `^`

# R uses math orders of operation: control with parentheses: ()

```

Other common mathematical operations use *functions*:

```{r computing2,echo=TRUE,results='markup'}
# calculate a square root using `sqrt`

# calculate the remainder using `%%`

# calculate a logarithm using `log`


```

What base does R assume you are using by default? How do we change the base?

```{r}


```

```{r}
exp(log(10))
```

R also recognizes scientific notation:
```{r}
5000000000 * 1000
5e9 * 1e3
```

# Assigning values to R objects

So far we have been mostly coding "interactively" - running a single command and 
looking at the answer - or we have been saving objects (like graphs saved as 
pictures) to our "working directory".

But what if we want to use the same object for later calculations?

A key action in R is to store values in the form of R objects, and to examine 
the value of R objects. (Add notes to help remind yourself what's going on!)

```{r assigning_values}
# (variable name) <- (value, or code to create a value/object)


# (variable name) = (value, or code to create a value/object)
# equal sign *can* be used for assignment but is not conventional

```
Notice how the values we saved weren't printed out?

```{r}
# In both the console and in R chunks, 
# running a variable name will print its value out


# We can also use the function `print`


```

Variable names have a few rules:

- can't start with a number
- case sensitive
- can contain some special characters: `.`, `_`

```{r}

```

You can over-write a variable once you've created it:
```{r}

```

You can see these values change in the "Environment" tab in Rstudio!


# Vectors in R

The most basic form of an R object is a vector. In fact, individual (scalar) 
values are vectors of length one. The variables we created earlier are all 
vectors!

We can concatenate values into a vector with `c()`.

```{r}
# Use `c()` with numbers

# Use `c()` with character strings

```

R Also has tools for defining repeatable patterns.

(DRY: "Don't Repeat Yourself")
```{r sequences}
# Use `:` to increment by integers

# Or use `seq` for more control

```

## 1) Numeric vectors
Also known as "double precision floating point"

```{r types_of_vectors1}


```

## 2) Integer vectors
Less common than numeric vectors, but it's good to know they exist since
sometimes you'll see errors related to the difference between numeric and 
integer vectors.
```{r types_of_vectors2}
# Create an integer by putting `L` at the end. 


```

Integers are not numbers!
```{r}
identical(5L, 5)
```

R Trivia: 'L' stands for 'long integer' (historical)

## 3) Character vector
Most common way of storing non-numeric data. 

```{r types_of_vectors3}

# create a character vector of your names

```

Character vectors must be wrapped in matching quotes (either " or ', but not a 
mix), or else R will interpret it as a variable.

```{r}
chars <- c('hi', 'hallo', "mother's", 'father\'s', 
   "She said, 'hi'", "He said, \"hi\"" )  
# the backslash is an "escape character"

```

Often we want to combine multiple character strings together:
```{r}
# use the function `paste0`

```


## 4) Logical vector (or boolean vector)
Logical values are either "TRUE" or "FALSE"

```{r types_of_vectors4}
# create a logical vector

```

Logical values are often created from numerical comparisons:
```{r}
vals <- c(1, 8, 10)

# Less than: <

# Less than or equal to: <=

# Greater than: >

# Greater than or equal to: >=

# Equal: ==

# Not equal: !=


```
Make sure you use double equal signs for equality!
Remember, single equal signs assigns:
```{r}
v1 <- 7
v2 <- 9
v1 == v2
v1 = v2
identical(v1, v2)
```

What if you combine data types?

```{r types_of_vectors5}
# create a combination vector:


# How do I know what datatype something is? function: `typeof`
```

# Working with indices and subsets

Use square brackets to refer to particular positions in a vector:

```{r indexing}
vals <- seq(2, 12, by = 2)
vals
vals[1]
```

Can you guess what these will do before running the code??
```{r indexing2}
vals
vals[3:5]
vals[c(1, 3)]
vals[-c(1, 3)]
```

We can also use logical values!
```{r}
which_indices <- c(rep(TRUE, 3), rep(FALSE, 2), TRUE)
which_indices
vals[which_indices]

vals == 6
vals[vals == 6]
vals[(vals %% 4) == 0]
```

# Comparing Values

We already saw numeric comparisons:
```{r comparisons}

vec1 <- c(1, 2, 3, 4, 5)
vec2 <- c(7, -4, 3, 4, 5.1)
vec2 < vec1
vec1 < vec2

vec1 == vec2
vec1 != vec2
```

We can also do other types of comparisons:
```{r}

## using 'or': |
vec1 >= 3 | vec1 < -7

## using 'and': &
vec1 >= 3 & vec1 %% 2 == 0 

```

What are the results of each of these?
```{r}
TRUE | FALSE
FALSE | FALSE
TRUE & FALSE
TRUE & TRUE
```

What answer do you expect here?
```{r}
"apple" < "banana"
```

# Breakout 1
In general, your answers to any questions should involve writing code to 
manipulate objects. Don't open excel or another calculator to answer them.

1) Create a vector of the squares of all numbers from 1-10. Save it to a
variable called `x`


```{r}
# your code goes here!

```

2) Use a logical expression to identify which elements of `x` are less than 20.
Save this logical vector as a variable called `y`.

```{r}


```

3) Use the vector `y` to subset x to return (a) all the elements less than 20 
and (b) all the elements greater than or equal to 20.

```{r}


```

4) How would you interpret the results of these expressions?
```{r}
sum(x)
mean(x)

sum(y)
mean(y)
```

# Reading Data From A File

```{r load_data}

#######################################################################
## IMPORTANT: read in the Gapminder dataset from disk;
## first make sure your working directory is the 'workbooks' directory:
getwd()
## if the result is not the 'workbooks' subdirectory of the bootcamp
## directory, navigate to this directory through the "Files" window,
## then click The "More (cogwheel)" > Set as working directory

## You can also set the working directly along the lines of this:
##
## setwd('/Users/jasminehughes/Documents/Intro_to_R/RforPharmacometricians2022')
##
## replace '/Users/jasminehughes/Documents/Intro_to_R/' with whatever directory 
## you put the bootcamp materials in; e.g. on Windows it might be something like
## 'C:\\Users\\sarah\\R_for_pharmacometricians'
##
## If you've done that correctly, then the next command reads
## in the dataset from the 'data' directory. In the next
## command R finds that directory relative to the current
## working directory.

gap <- read.csv(
  file.path("..", "data", "gapminder-FiveYearData.csv"),
  stringsAsFactors = FALSE
)



#######################################################################
```
Our data was imported as a dataframe. Dataframes are tabular: defined by rows and columns. When doing statistics & data analysis in R, you'll mostly be working with dataframes.

```{r looking_at_your_data}
## Look at part of the data
head(gap) # first few lines)
tail(gap) # last few lines
View(gap) # the whole shebang (avoid with very large datasets)
dim(gap) # dimensions
nrow(gap) # number of rows
summary(gap) # basic info about each column
```

# R is a functional language

* Operations are carried out with functions. Functions take objects as inputs 
  and return objects as outputs. 
* An analysis can be considered a pipeline of function calls, with output from a 
  function used later in a subsequent operation as input to another function.
* Functions themselves are objects: 

```{r functions1}
median
class(median)
```

```{r}
# Use `median` to calculate the median of a vector


```

Lets use this function to calculate the median life expectancy in the data set.

How do we refer to a column?

```{r}
# dollar notation: variable$column

# double square brackets: variable[["column"]]

```

Now we can calculate median life expectancy:

```{r functions2}

```

But wait, this data set has data from many years, what if we want to know the
median just for 2007?

```{r}
# calculate median for the year 2007:

```

* Functions generally take arguments, some of which are often optional.

* We can nest function calls: 

```{r, hist, fig.cap = ""}
random_data <- rnorm(1000, mean = 10, sd = 2)
hist(random_data)

hist(rnorm(1000, mean = 10, sd = 2))
```

# Getting help about a function

To get information about a function you know exists, use `help` or `?`, 
e.g., `?lm`. For information on a general topic, use `apropos`.

```{r eval = FALSE}
help(lm)
?lm

?log

apropos("mean")

```

# Converting between different types of objects

You can use the `as` family of functions.

```{r converting_between_types}
ints <- 1:10
as.character(ints)
as.numeric(c('3.7', '4.8'))
```

# Be Lazy

If you're starting to type something you've typed before, or the long name of an 
R object or function, STOP!  You likely don't need to type all of that.

- Tab completion
- Command history 
    * up/down arrows
    * Ctrl-{up arrow} or Command-{up arrow}
- RStudio: select a line or block for execution
- Write re-usable code! Put your code in a file and use `source()`. 
  Example: `source(file.path('..', 'example_scripts', 'Example_Rscript.R'))`. 

Or, create functions you can use within an analysis.

# Common issue: incomplete code

Sometimes you write some code that isn't a "complete sentence" and R gets
confused. 
```{r eval=FALSE}
(((9 + 2)) # can you spot the problem here?
```
The R console will show a little + sign, indicating it is expecting more.

Press escape to try again, or complete the "sentence" by adding the missing 
bracket.

# Breakout 2

In general, your answers to any questions should involve writing code to 
manipulate objects. For example, if I ask you to find the maximum GDP, 
do not scan through all the values and find it by eye. Use R to do the 
calculations and print results.

Some of these questions will require you to search for the right function.

**Hint**: use the function `apropos("key word")` to find something to use.


1) The variable x contains all the life expectancy data in the "gap" data frame 
we loaded earlier. Create a variable called 'y' that contains the mean life 
expectancy. Then, create a vector of length 2 that contains the minimum and 
maximum life expectancy and assign it to the value z.
```{r}
x <- gap$lifeExp
x

# y <- ...
# z <- ...

```

2) Use functions in R to round 'x' to two decimal places and to two 
significant digits.

```{r}
x

```

3) Create a vector of GDP per capita in units of Euros rather than dollars.

```{r}
exchange_rate <- 1.1

```

4) Create a boolean (TRUE/FALSE) vector indicating whether total country GDP is greater than 1 trillion dollars. When entering 1 trillion, use R's scientific notation.

```{r}

```
